---
title: "Great Whale Density Dependent Reproduction"
format:
  html:
    embed-resources: true
editor: visual
---

```{r}
#| label: initialize
#| include: false
library(here)
library(targets)

source(here("R", "plot.R"))
```

# Industrial whaling in the Southern Ocean

Industrial whaling harvested millions of whales, reducing some species' populations to \<1% of pre-whaling sizes.

## Catch trends

Data from the IWC. How do we properly attribute?

Catches over time. Color palette as in other reports (blue whale, fin whale, sperm whale = blue, green, orange). Notice shift to smaller species over time, as well as substantial decrease in effort during WWII.

```{r}
#| label: iwc_catches
#| echo: false

iwc_data <- tar_read("iwc_data")
ggplot(iwc_data, aes(catch_year, fill = species)) +
  geom_area(stat = "count") +
  scale_fill_manual(values = pal()) +
  labs(y = "Annual catches") +
  theme_classic() +
  theme(axis.title.x = element_blank(),
        legend.position = "none")
  
```

## Pregnancy trends

Pregnancy rates increased over time. Only years with \>0 females caught included.

```{r}
#| label: preg_rates
#| echo: false

preg_data <- tar_read("preg_data")

ggplot(preg_data, 
       aes(catch_year, preg_rate, size = f_catches, color = species)) +
  geom_point(alpha = 0.5) +
  geom_point(shape = 21) +
  scale_y_continuous("Pregnancy rate", labels = percent) +
  scale_color_manual(values = pal(), guide = "none") +
  scale_size_continuous(breaks = c(1, 5e3, 10e3)) +
  labs(size = "Females caught") +
  expand_limits(y = 0, size = 1) +
  theme_classic() +
  theme(axis.title.x = element_blank(),
        legend.justification = c(1, 1),
        legend.position = c(0.9, 0.9))
```

## Population trends

Data from [Christensen (2006) *UBC* *Fisheries Centre Research Reports*](http://doi.org/10.14288/1.0074757)*;* reconstructed population sizes of whale populations. Note: blue and fin populations from Southern Ocean but sperm populations global.

```{r}
#| label: pop_trends
#| echo: false

whale_pops <- tar_read("whale_pops")
ggplot(whale_pops, aes(catch_year, pop_size / 1000, color = species)) +
  geom_line() +
  scale_color_manual(values = pal()) +
  scale_y_continuous("Population (1000s)") +
  theme_classic() +
  theme(axis.title.x = element_blank(),
        legend.position = "none")
```

Population trends normalized as fraction of 20th century maximum.

```{r}
#| label: pop_trends_norm
#| echo: false

whale_pops_20th <- tar_read("whale_pops_20th")
ggplot(whale_pops_20th, aes(catch_year, pop_norm, color = species)) +
  geom_line() +
  scale_color_manual(values = pal()) +
  scale_y_continuous("Population (normalized)") +
  theme_classic() +
  theme(axis.title.x = element_blank(),
        legend.position = "none")
```

# Density dependent reproduction

Pregnancy rates were greater at low population sizes

```{r}
#| label: preg_pop
#| echo: false

preg_pop_data <- tar_read("preg_pop_data")
# In 1973 2 female blues were caught and both were pregnant, which makes the
# figure harder to interpret
ggplot(filter(preg_pop_data, preg_rate < 1),
       aes(pop_norm, preg_rate, size = f_catches, color = species)) +
  geom_point(alpha = 0.5) +
  geom_point(shape = 21) +
  scale_y_continuous("Pregnancy rate", 
                     labels = percent) +
  scale_color_manual(values = pal()) +
  labs(x = "Population (normalized)") +
  theme_classic() +
  theme(legend.position = "none")
```
