---
title: "Great Whale Density Dependent Reproduction"
format:
  html:
    embed-resources: true
editor: visual
---

```{r}
#| label: initialize
#| include: false
library(targets)

source(here("R", "plot.R"))
```

# Density dependence

Fit a logistic model to pregnancy rates w.r.t. population size and test two hypotheses. H1: great whales' pregnancy rates are negatively related to population size. H2: the magnitude of the negative relationship is stronger for fast-paced (i.e., baleen whales) than slow-paced (i.e., sperm whale) species. Limit model to years where population size \>=20% of maximum population size.

```{r}
#| label: logistic_model
#| echo: false

preg_pop_data <- tar_read("preg_pop_data")
preg_predictions <- tar_read("preg_predictions")

ggplot(preg_predictions, aes(pop_norm, preg_rate, color = species)) +
  geom_point(aes(size = f_catches), 
             filter(preg_pop_data, preg_rate < 1),
             shape = 21) +
  geom_ribbon(aes(fill = species, 
                  ymin = preg_rate_lwr, 
                  ymax = preg_rate_upr),
              alpha = 0.25,
              color = NA) +
  geom_line() +
  scale_color_manual(values = pal()) +
  scale_fill_manual(values = pal()) +
  labs(x = "Population size (normalized)",
       y = "Probability of pregnancy",
       color = "",
       fill = "",
       size = "Females caught (annual)") +
  theme_classic() +
  theme(legend.position = "bottom",
        legend.box = "vertical")
```

## H1: Do great whales exhibit density dependent reproduction?

All three species' pregnancy rates were negatively correlated with population size.

```{r}
#| label: h1
#| echo: false

preg_model <- tar_read("preg_model")
preg_coef <- coef(preg_model)[c(2, 5, 6)]
# Interaction terms are relative to the reference level
preg_coef <- preg_coef + c(0, preg_coef[1], preg_coef[1])
preg_pval <- coef(summary(preg_model))[, "Pr(>|z|)"][c(2, 5, 6)]

format_p <- function(p, digits) {
  template <- sprintf("<%%0.%df", as.integer(digits))
  ifelse(p < 10^-digits, 
         sprintf(template, 10^-digits),
         round(p, digits))
}

tibble(Species = c("Blue Whale", "Fin Whale", "Sperm Whale"),
       Coefficient = round(preg_coef, 3),
       `P-value` = format_p(preg_pval, 3)) %>% 
  kable() %>% 
  kable_styling(full_width = FALSE)

```

## H2: Is density dependent reproduction greater in faster-paced species?

Density dependence was significantly greater in fin whales than sperm whales. Also greater in blue whales than sperm whales but not significantly so, likely due to limited sperm whale sample size and large confidence intervals.

```{r}
#| label: h2
#| echo: false

preg_em <- emtrends(preg_model, "species", "pop_norm")
format_p <- function(p, digits) {
  template1 <- sprintf("p < %%0.%df", as.integer(digits))
  template2 <- sprintf("p = %%0.%df", as.integer(digits))
  ifelse(p < 10^-digits, 
         sprintf(template1, 10^-digits),
         sprintf(template2, round(p, digits)))
}
preg_pairs <- as_tibble(pairs(preg_em)) %>% 
  mutate(comparisons = str_split(contrast, " - "),
         annotations = format_p(p.value, 3))

plot(preg_em,
     xlab = "Coef") +
  geom_signif(
    comparisons = preg_pairs$comparisons,
    annotations = preg_pairs$annotations,
    step_increase = 0.1
  ) +
  expand_limits(x = 0) +
  labs(x = "Coefficient") +
  theme_classic() +
  theme(axis.title.y = element_blank())

```
