---
title: "Life History Report"
format:
  html:
    embed-resources: true
editor: visual
---

```{r}
#| label: initialize
#| include: false
library(here)
library(purrr)
library(targets)

source(here("R", "plot.R"))
```

# Scaling of life history traits

Quantify great whales' pace-of-life relative to mammals. Account for effects of body size and phylogeny. See [Jeschke and Kokko (2009) *Evolutionary Ecology*](https://doi.org/10.1007/s10682-008-9276-y), [Revell (2014) *Methods in Ecology and Evolution*](https://doi.org/10.1111/j.2041-210X.2011.00169.x), [Soriano-Redondo et al. (2020) *Nature Communications*](https://doi.org/10.1038/s41467-020-19256-0).

## Life history traits

First we need a database of life history traits, which we get from [Myhrvold et al. (2015) *Ecology*](https://doi.org/10.1890/15-0846R.1). Notice how life history traits vary substantially with body size. In this and other reports, orange represents Sperm Whales, green Fins, and blue Blues.

```{r}
#| label: life_history_plots
#| echo: false

mammal_lht <- tar_read("mammal_lht")
lht_cols <- c(
  "female_maturity_d", "gestation_d", "weaning_d", "longevity_y", 
  "litter_or_clutch_size_n", "inter_birth_interval_y"
)
lht_names <- c(
  "Female age at maturity (days)", "Gestation (days)", "Weaning (days)", 
  "Longevity (years)", "Fecundity (n offspring)", "Inter-birth interval (y)"
)
lht_plots <- map2(lht_cols, lht_names, life_history_plot, dat = mammal_lht)

plot_grid(plotlist = lht_plots, nrow = 2)
```

## Phylogeny

Next we need a tree. Use one from [Upham et al. (2019) *PLOS*](https://doi.org/10.1371/journal.pbio.3000494), downloaded from [VertLife](http://verlife.org/phylosubsets). Cetaceans highlighted in yellow.

TODO: "We constructed a single species-level phylogenetic consensus tree for birds using 1000 phylogenetic trees from the BirdTree project with the phytools package in R (ver. 3.5.2)." (Soriano-Redondo et al. 2020). Can we do something similar with the vertlife trees?

```{r}
#| label: phylogeny
#| echo: false

mammal_tr <- tar_read("mammal_tr")
phylogeny_plot(mammal_tr)
```

## Residuals

Calculate life history trait residuals while accounting for body size and phylogeny. See [Jeschke and Kokko (2009) *Evolutionary Ecology*](https://doi.org/10.1007/s10682-008-9276-y). Though [Freckleton (2002) *Journal of Animal Ecology*](https://doi.org/10.1046/j.1365-2656.2002.00618.x) warns against using residuals as data in subsequent analyses. They recommend multiple regression with body size as a predictor variable instead, although it's not immediately apparent to me how that would dovetail with dimensionality reduction.

## PCA

After reducing dimensionality of life history traits (using phylogenetic PCA) we are left with two axes: a primary axis corresponding to pace-of-life (maturity, longevity, post-natal investment, and inter-birth interval) and a secondary axis describing fecundity (large litters with brief pre-natal investment or vice versa). I find it interesting that fecundity is largely orthogonal to pace-of-life in this space.

```{r}
#| label: pca
#| echo: false

lht_pca <- tar_read("lht_pca")

biplot(lht_pca)
pol_plot(lht_pca, mammal_lht)
```
